---
title: "Variability Explorer Tool -- Version 2.0"
author: "Hossein Estiri"
date: "January, 2016"
---

## Introduction to the Application

This application has been built to help researchers (or any other potential data users) explore variability in Electronic Health Record (EHR) data. Multiple statistical procedures exist for detecting anomalies in data. Most of these procedures have been designed for time series data that, by definition, do not match characteristics of patient records stored daily in EHR systems. Other methods to detect variability in data might be considered too “statistical” for the average clinical researchers, and are expected to be conducted at the "study" level – not as an overall view of variability at database/warehouse levels. This tool provides an interactive visual solution to overview of variability in data from two lenses, an "exploratory" view and a "regression-based" view. 

Interactivity is a key attribute to VET's philosophy, through which the tool does not impose a specific point of variability to the users. Instead, the tool is designed to allow the users to explore variability and decide what repercussion might need to be implemented to the data when statistical analysis is going to be conducted.


## Inputs and Outputs

You can embed Shiny inputs and outputs in your document. Outputs are automatically updated whenever inputs change.  This demonstrates how a standard R plot can be made interactive by wrapping it in the Shiny `renderPlot` function. The `selectInput` and `sliderInput` functions create the input widgets used to drive the plot.

```{r, echo=FALSE}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

## Embedded Application

It's also possible to embed an entire Shiny application within an R Markdown document using the `shinyAppDir` function. This example embeds a Shiny application located in another directory:

```{r, echo=FALSE}
# shinyAppDir(
#   # system.file("app", package="shiny"),
#   options=list(
#     width="100%", height=550
#   )
# )
```

Note the use of the `height` parameter to determine how much vertical space the embedded application should occupy.

You can also use the `shinyApp` function to define an application inline rather then in an external directory.

In all of R code chunks above the `echo = FALSE` attribute is used. This is to prevent the R code within the chunk from rendering in the document alongside the Shiny components.



