---
title: "Variability Explorer Tool -- Version 2.0"
author: "Hossein Estiri"
date: "December, 2015"
---

This application is built to help researchers (or any other EHR data users) explore variability in Electronic Health Record data. Multiple statistical procedures exist to detect anomalies in data. Most of these procedures have been designed for time series data that, by definition, do not match characteristics of ordinary patient records from EHR systems. Other methods to detect variability in data might be considered too statistical for the average clinical researchers, and/or are expected to be conducted at a "study" level -- not as an overall veiw of variability at database/warehouse levels. This tool provides a solution to overview of variability in data from two lenses (where visual is inherent in both), a "exploratory" overview of and "regression-based" It is designed based on OMOP CDM, but can be easily adapted to other data models.

## Inputs and Outputs

You can embed Shiny inputs and outputs in your document. Outputs are automatically updated whenever inputs change.  This demonstrates how a standard R plot can be made interactive by wrapping it in the Shiny `renderPlot` function. The `selectInput` and `sliderInput` functions create the input widgets used to drive the plot.

```{r, echo=FALSE}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

## Embedded Application

It's also possible to embed an entire Shiny application within an R Markdown document using the `shinyAppDir` function. This example embeds a Shiny application located in another directory:

```{r, echo=FALSE}
# shinyAppDir(
#   # system.file("app", package="shiny"),
#   options=list(
#     width="100%", height=550
#   )
# )
```

Note the use of the `height` parameter to determine how much vertical space the embedded application should occupy.

You can also use the `shinyApp` function to define an application inline rather then in an external directory.

In all of R code chunks above the `echo = FALSE` attribute is used. This is to prevent the R code within the chunk from rendering in the document alongside the Shiny components.



